---
title: "Fitting Exercise (Week 8)"
---

## Data processing and exploration

Load packages and data
```{r}
#Load packages 
library(here)
library(tidyverse)
library(gtsummary)
library(knitr)
library(tidymodels) 
library(yardstick)
library(pROC)

#read csv data
drug <- read.csv(here("fitting-exercise", "Mavoglurant_A2121_nmpk.csv"))

#preview dataset
dplyr::glimpse(drug) 
```


Make a plot that shows a line for each individual, with DV on the y-axis and time on the x-axis. 
Stratify by dose.
```{r}
#spaghetti plot 
ggplot(drug, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  geom_line() +
  geom_point() +
  labs(title = "Spaghetti Plot of DV over TIME Stratified by DOSE",
       x = "TIME",
       y = "DV",
       color = "DOSE")
```

Very hard to tell the differences in the trajectory of DV across the DOSE groups,
try log-transforming DV and remake the plot.
```{r}
ggplot(drug, aes(x = TIME, y = log(DV), group = ID, color = factor(DOSE))) +
  geom_line() +
  geom_point() +
  labs(title = "Spaghetti Plot of log(DV) over TIME Stratified by DOSE",
       x = "TIME",
       y = "DV",
       color = "DOSE")
```
It looks like DV decreases faster over time among those from DOSE 25. 


Keeps only observations with OCC = 1.
```{r}
drug2<-drug[drug$OCC==1,]
```


Create a data frame of size 120 x 18 which excludes the observations with TIME = 0 
and contains variable Y (total amount of drug for each individual).
```{r}
#create a new data frame that contains only observations with TIME = 0 
drug3<-drug2[drug2$TIME == 0,]

#exlude observations with TIME = 0 then compute the sum of the DV variable for each individual using dplyr::summarize()
sum_dv <- drug2 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarize(Y = sum(DV))

#merge the two data frames above
merged_drug <- left_join(drug3, sum_dv, by = "ID")
```


Converts RACE, SEX and DOSE to factor variables and keeps variables Y,DOSE,AGE,SEX,RACE,WT,HT.
```{r}
merged_drug2<-merged_drug %>%
  mutate(RACE=factor(RACE),
         SEX=factor(SEX),
         DOSE=factor(DOSE)) %>%
  select(Y,DOSE,AGE,SEX,RACE,WT,HT)

#preview new data frame
dplyr::glimpse(merged_drug2) 
```


## EDA revisited

Make a descriptive table of the variables
```{r}
#get descriptive statistics of variables and create a table
table1<-gtsummary::tbl_summary(merged_drug2, statistic = list(
  all_continuous() ~ "{mean}/{sd}",
  all_categorical() ~ "{n} / {N} ({p}%)"
),)
knitr::kable(table1,caption = "Summary Table: Mean/SD or n/N(%)")
```
The majority of the cohort are of DOSE 25, SEX 1,and Race 1.


Make scatterplots between Y and continuous predictors (AGE, WT, HT)
```{r}
#Scatterplot: Y ~ AGE  
ggplot(merged_drug2, aes(AGE, Y)) + 
  geom_point()

#Scatterplot: Y ~ WT  
ggplot(merged_drug2, aes(WT, Y)) + 
  geom_point()

#Scatterplot: Y ~ HT  
ggplot(merged_drug2, aes(HT, Y)) + 
  geom_point()
```
No obvious linear associations observed between Y and AGE, WT, HT.


Make boxplots between Y and categorical predictors (DOSE, SEX, RACE)
```{r}
#Boxplot: Y ~ DOSE  
ggplot(merged_drug2, aes(factor(DOSE), Y)) + 
  geom_boxplot()

#Boxplot: Y ~ SEX  
ggplot(merged_drug2, aes(SEX, Y)) + 
  geom_boxplot()

#Boxplot: Y ~ RACE  
ggplot(merged_drug2, aes(RACE, Y)) + 
  geom_boxplot()
```
Individuals of different RACE groups or SEX groups have similar average value of Y.
There seems to be a positive association between Y and DOSE.


Make histograms for variables Y,AGE, WT, HT.
```{r}
#histogram for Y  
hist(merged_drug2$Y)

#histogram for AGE  
hist(merged_drug2$AGE)

#histogram for WT  
hist(merged_drug2$WT)

#histogram for HT.  
hist(merged_drug2$HT)
```
These variables have a generally normal distribution. No obvious outliers are observed. 


Make scatterplot matrix for variables Y,AGE, WT, HT to inspect correlations.
```{r}
pairs(merged_drug2[, c("Y", "AGE", "WT", "HT")], main = "Scatter Plot Matrix for Y,AGE, WT, HT")
```
WT and HT have relatively high correlation. 


Save the clean data frame locally
```{r}
saveRDS(merged_drug2, file =  here("fitting-exercise", "merged_drug2.rds"))
```

## Model fitting

Fit a linear model to the continuous outcome (Y) using the main predictor of interest DOSE,
and calculate RMSE and R-squared.
```{r}
#fit linear model: Y~DOSE
model1 <- lm(Y ~ DOSE, data = merged_drug2)
summary(model1)

# make predictions on the original data
predictions1 <- predict(model1, newdata = merged_drug2)

# Compute RMSE and R-squared
rmse_1 <- rmse_vec(merged_drug2$Y, predictions1)
rsquared_1 <- rsq_vec(merged_drug2$Y, predictions1)

# print RMSE and R-squared
cat("RMSE:", rmse_1, "\n")
cat("R-squared:", rsquared_1, "\n")
```
Y of DOSE 37.5 (p-value=0.002) and Y of DOSE 50 (p-value<.001) are significantly higher 
than Y of DOSE 25. RMSE is 666.3, and R-squared is 0.516. 


Fit a linear model to the continuous outcome (Y) using the all predictor,and calculate 
RMSE and R-squared.
```{r}
#fit linear model: Y~DOSE+AGE+SEX+RACE+WT+HT
model2 <- lm(Y ~ DOSE+AGE+SEX+RACE+WT+HT, data = merged_drug2)
summary(model2)

# make predictions on the original data
predictions2 <- predict(model2, newdata = merged_drug2)

# Compute RMSE and R-squared
rmse_2 <- rmse_vec(merged_drug2$Y, predictions2)
rsquared_2 <- rsq_vec(merged_drug2$Y, predictions2)

# print RMSE and R-squared
cat("RMSE:", rmse_2, "\n")
cat("R-squared:", rsquared_2, "\n")
```
After adjusting for covariates, Y of DOSE 37.5 (p-value=0.001) and Y of DOSE 50 
(p-value<.001) are significantly higher than Y of DOSE 25. One unit increase in WT (p-value<.001) 
is significantly associated with a decrease of 23.3 in Y. Other covariates are not 
significantly associated with Y. RMSE is 590.3, and R-squared is 0.62. This model 
has better fit than the model with only DOSE as predictor.



Fit a logistic model to the SEX using the main predictor DOSE,and calculate 
accuracy and ROC-AUC.
```{r}
# fit logistic model: SEX~DOSE
logistic_model <- logistic_reg(mode = "classification") %>%
  set_engine("glm") %>%
  set_mode("classification")

model3 <- logistic_model %>% 
  fit(SEX ~ DOSE, data = merged_drug2)

#calculate odds ratios
model_estimates <- tidy(model3, conf.int = TRUE)
model_estimates$odds_ratio <- exp(model_estimates$estimate)

# display model estimates and odds ratios
print(model_estimates[, c("term", "estimate", "conf.low", "conf.high", "odds_ratio","p.value")])


# Make predictions on the original data
prediction3 <- predict(model3, new_data = merged_drug2)
prediction3$SEX<-merged_drug2$SEX

# Compute accuracy
accuracy_value3 <- accuracy(prediction3, truth = SEX, estimate = .pred_class)

# Compute AUC
roc_curve3 <- roc(as.numeric(prediction3$SEX),as.numeric( prediction3$.pred_class))
auc_value3 <- auc(roc_curve3)

# Print accuracy and AUC
cat("Accuracy:", accuracy_value3$.estimate, "\n")
cat("AUC:", auc_value3, "\n")
```
There is no significant association between DOSE and SEX. The accuracy of the model 
is 0.867 and the AUC is 0.5. 



Fit a logistic model to the SEX using the main predictor DOSE,and calculate 
accuracy and ROC-AUC.
```{r}
# fit logistic model: SEX~DOSE
logistic_model <- logistic_reg(mode = "classification") %>%
  set_engine("glm") %>%
  set_mode("classification")

model4 <- logistic_model %>% 
  fit(SEX ~ DOSE + Y + AGE + HT + WT + RACE, data = merged_drug2)

#calculate odds ratios
model_estimates <- tidy(model4, conf.int = TRUE)
model_estimates$odds_ratio <- exp(model_estimates$estimate)

# display model estimates and odds ratios
print(model_estimates[, c("term", "estimate", "conf.low", "conf.high", "odds_ratio","p.value")])


# Make predictions on the original data
prediction4 <- predict(model4, new_data = merged_drug2)
prediction4$SEX<-merged_drug2$SEX

# Compute accuracy
accuracy_value4 <- accuracy(prediction4, truth = SEX, estimate = .pred_class)

# Compute AUC
roc_curve4 <- roc(as.numeric(prediction4$SEX),as.numeric( prediction4$.pred_class))
auc_value4 <- auc(roc_curve4)

# Print accuracy and AUC
cat("Accuracy:", accuracy_value4$.estimate, "\n")
cat("AUC:", auc_value4, "\n")
```
There is still no significant association between DOSE and SEX. Only HT is significantly
associated with SEX. The accuracy of the model is 0.958 and the AUC is 0.897. This model 
has better fit than the model with only DOSE as predictor. 
