---
title: "CDC Data Exercise (Week 5)"
subtitle: "Vaccine Hesitancy for COVID-19: County and local estimates"
editor: 
  markdown: 
    wrap: 72
---

The dataset of "Vaccine Hesitancy for COVID-19: County and local
estimates" from <https://data.cdc.gov/> was used for this exercise. This
dataset contains estimates of percentage of adults who had vaccine
hesitancy, Social Vulnerability Index, and percentages of different
races for all US counties. Because the complete dataset is over 150MB, I
queried and exported the data for only the counties in Georgia. Details
of the data can be found [here](#0){style="font-size: 11pt;"}.

## Load packages and data

```{r}
#Load packages 
library("readr")
library("here")
library("Hmisc")
library("tidyverse")
library("gtsummary")
library("knitr")

#read csv data
vaccine <- read.csv(here("cdcdata-exercise", "Vaccine_Hesitancy_for_COVID-19__County_and_local_estimates_20240206.csv"))

#preview dataset
dplyr::glimpse(vaccine) 

#create a new dataset that contains only the variables of interest
vaccine_2  <- 
  vaccine %>% 
    mutate(SVI.Category=factor(SVI.Category) ) %>%
    mutate(CVAC.Level.Of.Concern=factor(CVAC.Level.Of.Concern) ) %>%
    select(FIPS.Code, Estimated.hesitant, Percent.Hispanic, Percent.non.Hispanic.Black, Percent.non.Hispanic.Asian, SVI.Category , CVAC.Level.Of.Concern)
```

Now, we have created a new dataset which includes only the variables of
interest.

*FIPS.Code*: id for counties

*Estimated.hesitant*: Estimate of percentage of adults who describe
themselves as “probably not” or “definitely not” going to get a COVID-19
vaccine once one is available to them, based on national survey data

*Percent.Hispanic:* Percent of county population that is Hispanic

*Percent.non.Hispanic.Black:* Percent of county population that is
non-Hispanic Black

*Percent.non.Hispanic.Asian:* Percent of county population that is
non-Hispanic Asian

*SVI.Category*: SVI categorized as follows: Very Low (0.0-0.19), Low
(0.20-0.39); Moderate (0.40-0.59); High (0.60-0.79); Very High
(0.80-1.0)

*CVAC.Level.Of.Concern*: CVAC categorized as follows: Very Low
(0.0-0.19), Low (0.20-0.39); Moderate (0.40-0.59); High (0.60-0.79);
Very High (0.80-1.0)

## Explore and process data

```{r}
#inspect the summaries of the variables
describe(vaccine_2)
```

From the results above, none of the variables have missing entries or
unreasonable outliers. But we need to collapse some of the groups that
have very small numbers for the two categorical variables.

```{r}
#collapse small groups for SVI.Category and CVAC.Level.Of.Concern 
vaccine_2 <- vaccine_2 %>%
  mutate(SVI.Category = case_when(
    SVI.Category %in% c("Very Low Vulnerability", "Low Vulnerability") ~ "Low/Very Low",                            # combine Low and Very Low  
    TRUE ~ SVI.Category   )) %>% # leave other categories unchanged
    mutate(CVAC.Level.Of.Concern  = case_when(
    CVAC.Level.Of.Concern  %in% c("Very Low Concern", "Low Concern", "Moderate Concern") ~ "Very Low/Low/Moderate",   # combine Very Low, Low and Moderate
    TRUE ~ CVAC.Level.Of.Concern       # leave other categories unchanged
  ))
#check if recoding worked
describe(vaccine_2)
```

Now, SVI.Category has 4 categories and CVAC.Level.Of.Concern has 3
categories. The dataset is ready for analysis.

## Table and figures showing distribution of variables

```{r}
#get descriptive statistics of variables and create a table
ab<-gtsummary::tbl_summary(select(vaccine_2,-"FIPS.Code"), statistic = list(
  all_continuous() ~ "{mean}/{sd}",
  all_categorical() ~ "{n} / {N} ({p}%)"
),)
knitr::kable(ab,caption = "Summary Table: Mean/SD or n/N(%)")

#Histograms of Estimated.hesitant
hist(vaccine_2$Estimated.hesitant)
#Histograms of Percent.Hispanic
hist(vaccine_2$Percent.Hispanic)
#Histograms of Percent.non.Hispanic.Black
hist(vaccine_2$Percent.non.Hispanic.Black)
#Histograms of Percent.non.Hispanic.Asian
hist(vaccine_2$Percent.non.Hispanic.Asian	)
```

According to the histograms, we know none of the four continuous
variables have normal distribution.
