---
title: "CDC Data Exercise (Week 5)"
subtitle: "Vaccine Hesitancy for COVID-19: County and local estimates"
editor: 
  markdown: 
    wrap: 72
---

The dataset of "Vaccine Hesitancy for COVID-19: County and local
estimates" from <https://data.cdc.gov/> was used for this exercise. This
dataset contains estimates of percentage of adults who had vaccine
hesitancy, Social Vulnerability Index, and percentages of different
races for all US counties. Because the complete dataset is over 150MB, I
queried and exported the data for only the counties in Georgia. Details
of the data can be found [here](#0){style="font-size: 11pt;"}.

## Load packages and data

```{r}
#Load packages 
library("readr")
library("here")
library("Hmisc")
library("tidyverse")
library("gtsummary")
library("knitr")

#read csv data
vaccine <- read.csv(here("cdcdata-exercise", "Vaccine_Hesitancy_for_COVID-19__County_and_local_estimates_20240206.csv"))

#preview dataset
dplyr::glimpse(vaccine) 

#create a new dataset that contains only the variables of interest
vaccine_2  <- 
  vaccine %>% 
    mutate(SVI.Category=factor(SVI.Category) ) %>%
    mutate(CVAC.Level.Of.Concern=factor(CVAC.Level.Of.Concern) ) %>%
    select(FIPS.Code, Estimated.hesitant, Percent.Hispanic, Percent.non.Hispanic.Black, Percent.non.Hispanic.Asian, SVI.Category , CVAC.Level.Of.Concern)
```

Now, we have created a new dataset which includes only the variables of
interest.

*FIPS.Code*: id for counties

*Estimated.hesitant*: Estimate of percentage of adults who describe
themselves as "probably not" or "definitely not" going to get a COVID-19
vaccine once one is available to them, based on national survey data

*Percent.Hispanic:* Percent of county population that is Hispanic

*Percent.non.Hispanic.Black:* Percent of county population that is
non-Hispanic Black

*Percent.non.Hispanic.Asian:* Percent of county population that is
non-Hispanic Asian

*SVI.Category*: SVI categorized as follows: Very Low (0.0-0.19), Low
(0.20-0.39); Moderate (0.40-0.59); High (0.60-0.79); Very High
(0.80-1.0)

*CVAC.Level.Of.Concern*: CVAC categorized as follows: Very Low
(0.0-0.19), Low (0.20-0.39); Moderate (0.40-0.59); High (0.60-0.79);
Very High (0.80-1.0)

## Explore and process data

```{r}
#inspect the summaries of the variables
describe(vaccine_2)
```

From the results above, none of the variables have missing entries or
unreasonable outliers. But we need to collapse some of the groups that
have very small numbers for the two categorical variables.

```{r}
#collapse small groups for SVI.Category and CVAC.Level.Of.Concern 
vaccine_2 <- vaccine_2 %>%
  mutate(SVI.Category = case_when(
    SVI.Category %in% c("Very Low Vulnerability", "Low Vulnerability") ~ "Low/Very Low",                            # combine Low and Very Low  
    TRUE ~ SVI.Category   )) %>% # leave other categories unchanged
    mutate(CVAC.Level.Of.Concern  = case_when(
    CVAC.Level.Of.Concern  %in% c("Very Low Concern", "Low Concern", "Moderate Concern") ~ "Very Low/Low/Moderate",   # combine Very Low, Low and Moderate
    TRUE ~ CVAC.Level.Of.Concern       # leave other categories unchanged
  ))
#check if recoding worked
describe(vaccine_2)
```

Now, SVI.Category has 4 categories and CVAC.Level.Of.Concern has 3
categories. The dataset is ready for analysis.

## Table and figures showing distribution of variables

```{r}
#get descriptive statistics of variables and create a table
ab<-gtsummary::tbl_summary(select(vaccine_2,-"FIPS.Code"), statistic = list(
  all_continuous() ~ "{mean}/{sd}",
  all_categorical() ~ "{n} / {N} ({p}%)"
),)
knitr::kable(ab,caption = "Summary Table: Mean/SD or n/N(%)")

#Histograms of Estimated.hesitant
hist(vaccine_2$Estimated.hesitant)
#Histograms of Percent.Hispanic
hist(vaccine_2$Percent.Hispanic)
#Histograms of Percent.non.Hispanic.Black
hist(vaccine_2$Percent.non.Hispanic.Black)
#Histograms of Percent.non.Hispanic.Asian
hist(vaccine_2$Percent.non.Hispanic.Asian	)
```

According to the histograms, we know none of the four continuous
variables have normal distribution.

## This section contributed by Patrick Kaggwa

```{r}
#Looking at the columns of vaccine2 before creating a synthetic dataset 
colnames(vaccine_2)
```

After undertanding the description of the above dataframe, I used
ChatGPT to create a new dataframe called vaccine_2pk with the same
structure as vaccine_2.

### ChatGPT prompt 

Create a dataset that should consist of 159 observations and include the
following variables: FIPS codes, Estimated hesitant values, Percent
Hispanic values, Percent non-Hispanic Black values, Percent non-Hispanic
Asian values, SVI Category, and CVAC Level of Concern. To ensure
reproducibility, set the seed to 123 before generating random values.
The FIPS codes should range from 13001 to 13159. The Estimated hesitant
values should be uniformly distributed between 0.1049 and 0.188. Percent
Hispanic values should range uniformly between 0.0014 and 0.3511.
Percent non-Hispanic Black values should range uniformly between 0.0025
and 0.7313. Percent non-Hispanic Asian values should range uniformly
between 0 and 0.1291. The SVI Category should be sampled with
replacement from 'High Vulnerability', 'Low/Very Low', 'Moderate
Vulnerability', and 'Very High Vulnerability', with probabilities of
31/159, 25/159, 21/159, and 82/159 respectively. Similarly, the CVAC
Level of Concern should be sampled with replacement from 'High Concern',
'Very High Concern', and 'Very Low/Low/Moderate', with probabilities of
50/159, 77/159, and 32/159 respectively. Create a dataframe with these
variables named 'vaccine_2pk.

```{r}

# Set seed for reproducibility
set.seed(123)

# Generate FIPS codes
FIPS.Code <- seq(13001, 13159)

# Generate Estimated hesitant values
Estimated.hesitant <- runif(159, min = 0.1049, max = 0.188)

# Generate Percent Hispanic values
Percent.Hispanic <- runif(159, min = 0.0014, max = 0.3511)

# Generate Percent non-Hispanic Black values
Percent.non.Hispanic.Black <- runif(159, min = 0.0025, max = 0.7313)

# Generate Percent non-Hispanic Asian values
Percent.non.Hispanic.Asian <- runif(159, min = 0, max = 0.1291)

# Generate SVI Category with exact distribution
SVI.Category <- sample(c("High Vulnerability", "Low/Very Low", "Moderate Vulnerability", "Very High Vulnerability"),
                        159, replace = TRUE, prob = c(31/159, 25/159, 21/159, 82/159))

# Generate CVAC Level of Concern
CVAC.Level.Of.Concern <- sample(c("High Concern", "Very High Concern", "Very Low/Low/Moderate"),
                                159, replace = TRUE, prob = c(50/159, 77/159, 32/159))

# Create the dataframe
vaccine_2pk <- data.frame(
  FIPS.Code = FIPS.Code,
  Estimated.hesitant = Estimated.hesitant,
  Percent.Hispanic = Percent.Hispanic,
  Percent.non.Hispanic.Black = Percent.non.Hispanic.Black,
  Percent.non.Hispanic.Asian = Percent.non.Hispanic.Asian,
  SVI.Category = SVI.Category,
  CVAC.Level.Of.Concern = CVAC.Level.Of.Concern
)

# Print summary statistics for each variable
summary(vaccine_2pk)



```

```{r}
# I will desecribe the da
describe(vaccine_2pk)
```

```{r}
#get descriptive statistics of variables and create a table
ab<-gtsummary::tbl_summary(select(vaccine_2pk,-"FIPS.Code"), statistic = list(
  all_continuous() ~ "{mean}/{sd}",
  all_categorical() ~ "{n} / {N} ({p}%)"
),)
knitr::kable(ab,caption = "Summary Table: Mean/SD or n/N(%)")

#Histograms of Estimated.hesitant
hist(vaccine_2pk$Estimated.hesitant)
#Histograms of Percent.Hispanic
hist(vaccine_2pk$Percent.Hispanic)
#Histograms of Percent.non.Hispanic.Black
hist(vaccine_2pk$Percent.non.Hispanic.Black)
#Histograms of Percent.non.Hispanic.Asian
hist(vaccine_2pk$Percent.non.Hispanic.Asian	)
```

The histograms show less skewed distribution but at the same time they are not
normally distributed. This is consistent with the histograms using dataframe vaccine_2
